---
title: "Exercise Key"
subtitle: "Logistic Regression"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/aaa_ExcKeys/aaa_rmd/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_Logistic_Ruffe1")
```
```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
source("../../../rhelpers/knitr_setup.R")
library(captioner)
tbls <- captioner(prefix="Table")
figs <- captioner(prefix="Figure")
figs(name="fitplot","Fitted plot for the logistic regression of whether or not a larval Ruffe had consumed a <i>Daphnia</i> and the length of the Ruffe.")
tbls(name="sum","Summary of the coefficients from the logistic regression of whether or not a larval Ruffe had consumed a <i>Daphnia</i> and the length of the Ruffe. ")
tbls(name="ci","Confidence intervals for the coefficients from the logistic regression of whether or not a larval Ruffe had consumed a <i>Daphnia</i> and the length of the Ruffe. ")
```
```{r analysis, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(NCStats)
df <- read.csv("https://raw.githubusercontent.com/droglenc/NCData/master/RuffeLarvalDiet.csv")
df <- filterD(df,loc=="Allouez")
glm1 <- glm(o.daph~len,data=df,family=binomial)
cf1 <- coef(glm1)
sum1 <- summary(glm1)
ci1 <- confint(glm1)
predProb <- function(x,alpha,beta) exp(alpha+beta*x)/(1+exp(alpha+beta*x))
p6 <- predProb(6,cf1[[1]],cf1[[2]])
bc2 <- bootCase(glm1)      # bootstrapping, be patient!
p6bc <- predProb(6,bc2[,1],bc2[,2])
p6ci <- quantile(p6bc,c(0.025,0.975),na.rm=TRUE)
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
  <li>Make sure to use confidence intervals where you can.</li>
  <li>I find it easier to talk about odds that are greater than 1. Thus, you will see that I "flipped" the odds below so I could talk about how many more times greater it was that the Ruffe had not consumed a *Daphnia* than it had.<li>
  <li>Note in question 3 that you cannot describe the relationship between the PROBABILITY and Ruffe length from the slope or the back-transformed slope; rather you must describe the shape of the curve in the fitplot. The slope tells how the log(odds) change (additively) with an increase in length and the back-transformed slope tells how the odds change (multiplicatively) with an increase in length.<li>
</ul>
</div>

----

## Ruffe Feeding

1. The fitted-line plot (`r figs("fitplot",display="cite")`) suggests that the logistic regression model fits the proportions of larval Ruffe that had consumed a *Daphnia* fairly well as indicated by the relative closeness of the modeled line to the observed proportions (i.e., blue pluses).
1. There is a significant relationship between the probability of consuming a *Daphnia* and the length of the Ruffe as indicated by a p-value for the slope of the logistic regression (`r kPvalue(sum1$coefficients[2,4],digits=4,latex=FALSE)`; `r tbls("sum",display="cite")`) that is less than 0.05.
1. The relationship between the probability of having consumed a *Daphnia* and the length of the Ruffe cannot be described with any given number and must be described from `r figs("fitplot",display="cite")`.  From `r figs("fitplot",display="cite")` it is seen that this probability is "low" until about 6 mm, rises sharply to about 10 mm, and then stays high above 10 mm. Additionally, the odds of consuming a *Daphnia* are between `r formatC(exp(ci1[2,1]),format="f",digits=4)` and `r formatC(exp(ci1[2,2]),format="f",digits=4)` times greater for each increase of 1 mm in larval Ruffe length (back-transformed from `r tbls("ci",display="cite")`).
1. The odds that a 6-mm long larval Ruffe had consumed a *Daphnia* is `r formatC(p6/(1-p6),format="f",digits=4)`, which means that the probability that a 6-mm long larval Ruffe had consumed a *Daphnia* is only `r formatC(p6/(1-p6),format="f",digits=4)` times the probability that it had not consumed a *Daphnia*. This indicates that the 6-mm long larval Ruffe was `r formatC((1-p6)/p6,format="f",digits=1)` (=$\frac{1}{`r formatC(p6/(1-p6),format="f",digits=4)`}$) times more likely to NOT have consumed a *Daphnia* then it was to have consumed a *Daphnia*.
1. The probability that a 6-mm long larval Ruffe had consumed a *Daphnia* is `r formatC(p6,format="f",digits=4)`, with a 95% confidence interval from `r formatC(p6ci[1],format="f",digits=4)` to `r formatC(p6ci[2],format="f",digits=4)`. This means that approximately one out of every `r formatC(1/p6,format="f",digits=1)` (95% CI: `r formatC(1/p6ci[2],format="f",digits=4)`, `r formatC(1/p6ci[1],format="f",digits=4)`) 6-mm long larval Ruffe will have consumed a *Daphnia*.


```{r echo=FALSE, fig.width=3.95}
par(mar=c(3.05,3.05,0.65,1.55),mgp=c(2.1,0.3,0),tcl=-0.2,
    las=1,cex.lab=0.95,cex.axis=0.9)
fitPlot(glm1,xlab="Length (mm)",ylab="Daphnia in Stomach?",breaks=seq(3.5,17.5,1))
```

`r figs("fitplot")`

`r tbls("sum")`

```{r echo=FALSE, background="white"}
sum1$coefficients
```


`r tbls("ci")`

```{r echo=FALSE, background="white"}
ci1
```

R Appendix.
```{r eval=FALSE, prompt=FALSE}
<<analysis>>
fitPlot(glm1,xlab="Length (mm)",ylab="Daphnia in Stomach?",breaks=seq(3.5,17.5,1))
```

----
