---
title: "Exercise Key"
subtitle: "1-Way ANOVA"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/aaa_ExcKeys/aaa_rmd/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_Anova-1way_Tables2")
```
```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
library(NCStats)
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
<li>$MS_{Total}$ is not usually shown in an ANOVA table). I put it in the table here because that is one of the given values (see the notes below) and it makes it easier to see how the rest of the table is completed.</li>
</ul>
</div>

----

## ANOVA Table 2

Source     | df | SS | MS | F  | p-value
:----------|---:|---:|---:|---:|--------:
**Among**  |  5 | 887.05 | 177.41 | 5.25 | `r kPvalue(distrib(5.25,distrib="f",df1=5,df2=48,lower.tail=FALSE),include.p=FALSE,latex=FALSE,digits=3)` |
**Within** | 48 | 1621.44 | 33.78 | - | -
**Total**  | 53 | 2508.49 | *47.33** | - | -

1. See completed table above. Note the following
    * $MS_{total}$ and $MS_{within}$ are given by definition (the variance of individuals ignoring groups is $MS_{total}$ whereas the pooled (or combined) variance of individuals considering groups is $MS_{within}$).
    * The $df$ are found by knowing the number of groups ($I=6$) and total number of individuals ($n=54$).
    * $SS_{within}$ and $SS_{total}$ are obtained with $MS*df$.
    * $SS_{among}$ is from $SS_{total}-SS_{within}$.
    * $MS_{among}$ is from $SS_{among}$ divided by $df_{among}$.
    * $F$ is from $MS_{among}$ divided by $df_{among}$.
    * The p-value is computed with `distrib(5.25,distrib="f",df1=5,df2=48,lower.tail=FALSE)`.
1. Yes, there is a difference among the group means because the p-value is less than 0.05.

----
