---
title: "Exercise Key"
subtitle: "Indicator Variable Regression"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/aaa_ExcKeys/aaa_rmd/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_IVR_Pronghorns")
```
```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
library(NCStats)
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
  <li>The indicator variable is named after the non-reference group.</li>
  <li>The variables are entered into the ultimate full model in this order -- covariate, indicator variable, and interaction variable.</li>
</ul>
</div>

----

## Growth of Pronghorn Antelopes

1. The indicator variable is named "TAME" where TAME=1 if the pronghorn is in the "tame" group and TAME=0 if it is in the "diet-curtailed" group.

1. The ultimate full model is &mu;<sub>WEIGHT</sub> = &alpha; + &beta;TIME + &delta;<sub>1</sub>TAME + &gamma;<sub>1</sub>TAME*TIME.

1. The submodels for both groups are below
    * Diet-curtailed: &alpha; + &beta;TIME
    * Tame: (&alpha;+&delta;<sub>1</sub>) + (&beta;+&gamma;<sub>1</sub>)TIME

1. Interpretations of the coefficients are below.
    * &alpha; is the intercept of diet-curtailed (reference) group
    * &beta; is the slope of diet-curtailed (reference) group
    * &delta;<sub>1</sub> is the difference in intercept of tame and diet-curtailed groups (i.e., tame-dietcurtailed)
    * &gamma;<sub>1</sub> is the difference in slopes of tame and diet-curtailed groups (i.e., tame-dietcurtailed)

1. The models for the "parallel lines test" are below.
    * H<sub>0</sub>: &mu;<sub>WEIGHT</sub> = &alpha; + &beta;TIME + &delta;<sub>1</sub>TAME
    * H<sub>A</sub>: &mu;<sub>WEIGHT</sub> = &alpha; + &beta;TIME + &delta;<sub>1</sub>TAME + &gamma;<sub>1</sub>TAME*TIME

1. The models for the "intercepts test" (assuming parallel lines) are below.
    * H<sub>0</sub>: &mu;<sub>WEIGHT</sub> = &alpha; + &beta;TIME
    * H<sub>A</sub>: &mu;<sub>WEIGHT</sub> = &alpha; + &beta;TIME + &delta;<sub>1</sub>TAME

----
