---
title: "Exercise Key"
subtitle: "1-Way ANOVA"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/aaa_ExcKeys/aaa_rmd/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_Anova-1way_Tables1")
```
```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
library(NCStats)
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
<li>$MS_{Total}$ is not usually shown in an ANOVA table but it is an important value (i.e., the overall variance ignoring the groups).</li>
</ul>
</div>

----

## ANOVA Table 1

Source     | df | SS | MS | F  | p-value
:----------|---:|---:|---:|---:|--------:
**Among**  |  **7** | **35819** | **5117** | **3.5** | **`r kPvalue(distrib(3.5,distrib="f",df1=7,df2=24,lower.tail=FALSE),include.p=FALSE,latex=FALSE,digits=3)`** |
**Within** | 24 | 35088 | **1462** | - | -
**Total**  | 31 | 70907 | - | - | -

1. See completed table above. Note the following
    * $df_{among}$ and $SS_{among}$ came from subtracting within values from total values.
    * $MS_{among}$ came from $SS_{among}$ divided by $df_{among}$.
    * $F$ is then $MS_{among}$ divided by $MS_{within}$.
    * The p-value came from `distrib(3.5,distrib="f",df1=7,df2=24,lower.tail=FALSE)`.
1. The number of groups = 7+1 = 8 (i.e., $df_{among}$+1).
1. The number of individuals = 31+1 = 32 (i.e., $df_{total}$+1).
1. The variability among individuals within groups is $s_{p}^{s}$ = $MS_{within}$ = 1462.
1. The variability among individuals ignoring groups is $s^{2}$ = $MS_{total}$ = $\frac{70907}{31}$ = 2287.3.
1. Yes, there is a difference among the group means because the p-value is less than 0.05.

----
