---
title: "Exercise Key"
subtitle: "One-Way ANOVA"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/aaa_ExcKeys/aaa_rmd/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_Anova-1way_Tables3")
```
```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
library(NCStats)
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
<li>MS<sub>Total</sub> is not usually shown in an ANOVA table but it is an important value (i.e., the overall variance ignoring the groups).</li>
<li>When you see a question about a "variance" then it will be answered with an "MS". Questions about variance of individuals will NOT be MS<sub>Among</sub>. Variances ignoring groups uses the simple model and will thus be MS<sub>Total</sub>.</li>
</ul>
</div>

----

## ANOVA Table 3

Source     | df | SS | MS | F  | p-value
:----------|---:|---:|---:|---:|--------:
**Among**  | **3** | 17.25 | **5.75** | 1.26 | **`r kPvalue(distrib(1.26,distrib="f",df1=3,df2=20,lower.tail=FALSE),include.p=FALSE,latex=FALSE,digits=3)`**
**Within** | **20** | **91.20** | 4.56 | - | -
**Total**  | 23 | **108.45** | -    | - | -

1. See completed table above. Note the following
    * MS<sub>among</sub> is from multiplying F and MS<sub>within</sub>.
    * df<sub>among</sub> is then SS<sub>among</sub> divided by MS<sub>among</sub>.
    * df<sub>within</sub> is then df<sub>total</sub>-df<sub>among</sub>.
    * SS<sub>within</sub> is then MS<sub>within</sub> times df<sub>within</sub>.
    * SS<sub>total</sub> is then SS<sub>among</sub> + SS<sub>within</sub>.
    * The p-value is from `distrib(1.26,distrib="f",df1=3,df2=20,lower.tail=FALSE)`.
1. The number of groups = 3+1 = 4 (i.e., $df<sub>among</sub>+1).
1. The number of individuals = 23+1 = 24 (i.e., $df<sub>total</sub>+1).
1. The variability among individuals within groups is s<sub>p</sub><sup>2</sup> = $MS<sub>within</sub> = 4.56.
1. The variabilty among individuals ignoring groups is s>sup>2</sup> = MS<sub>total</sub> = $\frac{108.45}{23}$ = 4.72.
1. No there is not a difference among the group means because the p-value is greater than 0.05.

----
