---
title: "Model Types & Methods"
description: |
  Introduction to linear model types covered during this course.
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(NCStats)
library(ggplot2)
knitr::opts_chunk$set(echo=FALSE,message=FALSE,comment="#R>  ",
                      fig.align='center',fig.width=3.5,fig.height=3.5)
data(Mirex)
Mirex <- Mirex %>%
  dplyr::mutate(year=factor(year),
                exceeds_limit=factor(ifelse(mirex>0.1,"Yes","No")))
```

During this course we will examine a variety of models called either *general linear* or *generalIZED linear* models. General linear models have a quantitative [response variable](https://deepai.org/machine-learning-glossary-and-terms/response-variable) and generally assume that the "errors" around the model follow a normal distribution. General linear models that we will discuss are a **One-Way ANOVA**, **Two-WAY ANOVA**, **Simple Linear Regression**, and **Indicator Variable Regression**. GeneralIZED linear models do not require a quantitative response variable nor "errors" that are normally distributed. Thus, generalIZED linear models are more flexible than general linear models. The only generalIZED linear model that we will encounter in this course is **Logistic Regression**, but the chi-square test from your introductory statistics course can also be cast as a generalIZED linear model.

<aside>
ANOVA is short for ANalysis Of VAriance
</aside>

Both general and generalIZED linear models can have a single explanatory variable that can be either quantitative or categorical, or multiple explanatory variables that can be all quantitative, all categorical, or a mixture of both quantitative and categorical. Ultimately, there can be several explanatory variables in a model, but we will only consider one or two explanatory variables in this course.

<aside>
All models covered in this course will have **only one** response variable
</aside>

&nbsp;

# Distinguishing Methods
The five methods that will be covered in this course can be distinguished by considering only the type of response variable and the types and number of explanatory variables (see table below). Thus, you will want to review [variable types](http://derekogle.com/NCMTH107/modules/FoundationalDefns.html) and [definitions and distinctions of response and explanatory variables](http://derekogle.com/NCMTH107/modules/DataProduction.html) from your introductory statistics course.

| Response     | Explanatory                            | Linear Model                  |
|--------------|----------------------------------------|-------------------------------|
| Quantitative | Categorical (only one)                 | One-Way ANOVA                 |
| Quantitative | Categorical (two)                      | Two-Way ANOVA                 |
| Quantitative | Quantitative (only one)                | Simple Linear Regression      |
| Quantitative | Quantitative (one) & Categorical (one) | Indicator Variable Regression |
| Binomial     | Quantitative (or Both)                 | (Binary) Logistic Regression  |

<aside>
How to distinguish among methods covered in this course
</aside>

&nbsp;

# Method Purposes
As seen above, each method uses different types of data. Not surprisingly then, each method is used to test different hypotheses or has a different analytical purpose. These purposes will be discussed in detail in subsequent modules. However, the major objective of each method is explained briefly below (in the order that we will cover them).

Each example uses a data set that contains data about [mirex](https://en.wikipedia.org/wiki/Mirex) concentrations (`mirex`) for two species of salmon (`species`) captured in six years between 1977 and 1999 (`year`) in Lake Ontario. The weight of each fish (`weight`) and whether or not the mirex concentration exceeded the EPA limit of 0.1 mg/kg (`exceeds_limit`) were also recorded.

&nbsp;

A **one-way ANOVA** is used to determine if the means of the quantitative response variable (`mirex`) differ among two or more groups defined by a single categorical variable (e.g., `year`).

```{r}
ggplot(data=Mirex,mapping=aes(x=year,y=mirex)) +
  geom_point(alpha=0.15) +
  stat_summary(fun.data=mean_cl_normal,geom="errorbar",width=0.25,size=1) +
  stat_summary(fun=mean,geom="point",shape=15,size=2) +
  labs(y="Mirex Concentration (mg/kg)",x="Sample Year") +
  theme_NCStats()
```

&nbsp;

A **two-way ANOVA** is used to determine if the means of the quantitative response variable (`mirex`) differ among groups of one categorical variable (e.g., `year`), among groups of another categorical variable (e.g., `species`), or by the interaction between the two categorical variables.


```{r}
pd <- position_dodge(0.25)
ggplot(data=Mirex,mapping=aes(x=year,y=mirex,color=species)) +
  geom_point(alpha=0.25,position=pd) +
  stat_summary(fun.data=mean_cl_normal,geom="errorbar",width=0.25,size=1,position=pd) +
  stat_summary(fun=mean,geom="point",shape=15,size=2,position=pd) +
  labs(y="Mirex Concentration (mg/kg)",x="Sample Year") +
  theme_NCStats() +
  theme(legend.position=c(1,1),legend.justification=c(1,1.02))
```

&nbsp;

A **simple linear regression** is used to determine if there is a relationship between the quantitative response variable (e.g., `mirex`) and a single quantitative explanatory variable (e.g., `weight`).

```{r}
ggplot(data=Mirex,mapping=aes(x=weight,y=mirex)) +
  geom_point(alpha=0.25) +
  geom_smooth(method="lm",se=FALSE) +
  labs(y="Mirex Concentration (mg/kg)",x="Weight (kg)") +
  theme_NCStats()
```

&nbsp;

An **indicator variable regression** is used to determine if the relationship between a quantitative response (e.g., `mirex`) and a quantitative explanatory variable (e.g., `weight`) differs between two or more groups defined by a categorical explanatory variable (e.g., `species`). This will look like two (or more) simple linear regressions are being compared.

```{r}
ggplot(data=Mirex,mapping=aes(x=weight,y=mirex,color=species)) +
  geom_point(alpha=0.25) +
  geom_smooth(method="lm",se=FALSE) +
  labs(y="Mirex Concentration (mg/kg)",x="Weight (kg)") +
  theme_NCStats() +
  theme(legend.position=c(0,1),legend.justification=c(-0.02,1.02))
```

&nbsp;

A **logistic regression** is used to determine if there is a relationship between the probability of "success" for a binary^[Binary means there are only two categories -- generically "success" and "failure".] categorical response variable (e.g., `exceeds_limit`) and the quantitative explanatory variable (e.g., `weight`).

```{r}
ggplot(data=dplyr::filter(Mirex,species=="coho"),
       mapping=aes(x=weight,y=as.numeric(exceeds_limit)-1)) +
  geom_point(alpha=0.25) +
  geom_smooth(method="glm",se=FALSE,
              method.args=list(family="binomial")) +
  labs(y="Prob. of Exceeding 0.1 mg/kg Mirex",x="Weight (kg)") +
  theme_NCStats()
```

&nbsp;

From these examples it should be apparent that "ANOVAs" are about **comparing means** among groups and will look like means (usually with confidence intervals) plotted as points for each group. In contrast "regressions" are about exploring **relationships** and will look like a line or a curve when plotted.

<aside>
ANOVAs compare means; regressions examine relationships.
</aside>

# Corrections {.appendix}
If you see any errors (code, typographical, or logical) please bring these to [my attention](mailto:dogle@northland.edu).
