---
title: "Exercise Key"
subtitle: "Logistic Regression Foundations"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/ce/aaaKeys/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_LogRegFoundations_CE1")
```

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(NCStats)
source("../../../rhelpers/knitr_setup.R")
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
  <li>Use complete sentences to answer questions.</li>
</ul>
</div>

----

# Fair Coin

1. The probability of flipping a head is $\frac{1}{2}$=0.5.
1. The odds of flipping a head is $\frac{0.5}{0.5}$=1.
1. The odds of flipping a tail is $\frac{0.5}{0.5}$=1.
1. the log odds of flipping a head is log(1)=0.

&nbsp;

# Fair Die
```{r echo=FALSE}
p6 <- round(1/6,3)
odds6 <- round((1/6)/(5/6),3)
lodds6 <- round(log((1/6)/(5/6)),3)
pn6 <- 1-p6
oddsn6 <- 1/odds6
loddsn6 <- log(oddsn6)
```

1. The probability of rolling a six is $\frac{1}{6}$=`r p6`.
1. The odds of rolling a six is $\frac{`r p6`}{1-`r p6`}$=`r odds6`.
1. The log odds of rolling a is log(`r odds6`)=`r lodds6`.
1. The probability of *not* rolling a six is $\frac{5}{6}$=`r pn6`.
1. The odds of *not* rolling a six is $\frac{`r pn6`}{1-`r pn6`}$=`r oddsn6`.
1. The log odds of *not* rolling a is log(`r oddsn6`)=`r loddsn6`.
1. The probability of rolling a four, five, or six is $\frac{3}{6}$=0.5.
1. The odds of rolling a four, five, or six is $\frac{0.5}{0.5}$=1.
1. The log odds of rolling a four, five, or six is log(1)=0.
1. The odds are 1 when the probability is 0.5.
1. The odds are between 0 and 1 when the probability is <0.5.
1. The odds are >1 when the probability is >0.5.
1. The log odds are 0 when the probability is 0.5.
1. The log odds are negative when the probability is <0.5.
1. The log odds are positive when the probability is >0.5.

&nbsp;

# Marketability of Cucumbers
```{r echo=FALSE}
alpha <- 6.430
beta <- -0.330
DOS <- c(14,15)
cuc.lodds <- alpha+beta*DOS
cuc.odds <- exp(cuc.lodds)
cuc.p <- cuc.odds/(1+cuc.odds)
beta2 <- diff(cuc.lodds)
ebeta2 <- tail(cuc.odds,1)/head(cuc.odds,1)
ebeta <- exp(beta)
```

1. The log odds for the sale of a cucumber after 14 days in storage is `r alpha``r beta`&times;`r DOS[1]`=`r round(cuc.lodds[1],3)`.
1. The odds for the sale of a cucumber after 14 days in storage is e<sup>`r round(cuc.lodds[1],3)`</sup>=`r round(cuc.odds[1],3)`.
1. The probability for the sale of a cucumber after 14 days in storage is $\frac{`r round(cuc.odds[1],3)`}{1+`r round(cuc.odds[1],3)`}$=`r round(cuc.p[1],3)`.
1. The log odds for the sale of a cucumber after 15 days in storage is `r alpha``r beta`&times;`r DOS[2]`=`r round(cuc.lodds[2],3)`.
1. The difference of log odds is `r round(cuc.lodds[2],3)`-`r round(cuc.lodds[1],3)`=`r round(beta2,3)`. This is the same as the slope (&beta;); i.e., the **difference in log odds** after a one unit increase in the explanatory variable.
1. The odds for the sale of a cucumber after 15 days in storage is e<sup>`r round(cuc.lodds[2],3)`</sup>=`r round(cuc.odds[2],3)`.
1. The ratio of odds is $\frac{`r round(cuc.odds[2],3)`}{`r round(cuc.odds[1],3)`}$=`r round(ebeta2,3)`. This is the same as the exponentiated slope (e<sup>&beta;</sup>=e<sup>`r beta`</sup>=`r round(ebeta,3)`); i.e., the **ratio of odds** after a one unit increase in the explanatory variable.
