---
title: "Exercise Key"
subtitle: "IVR Variables"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/ce/aaaKeys/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_IVRVariables_CE1")
```

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(NCStats)
source("../../../rhelpers/knitr_setup.R")
```

----

# Nurse Wages

1. The response variable is **WAGE**.
1. The covariate is months of **EXPERience**.
1. The factor is **gender** (female and male).
1. As there are two groups in this analysis and females comes before males, I created the following indicator variable.
    * MALE=1 if "male", 0 otherwise (i.e., "female")
1. The very simple table connecting the groups to the indicator variables codes is as follows:
```{r echo=FALSE}
data.frame(group=c("female","male"),MALE=c(0,1)) %>%
  knitr::kable(booktabs=TRUE,row.names=FALSE) %>%
  kableExtra::kable_classic(full_width=FALSE,html_font="Cambria")
```

<ol start="6">
<li>The only interaction variables would be between months of EXPERience and MALE; thus, experience:FEM.</li>
<ol>

&nbsp;

# Turtle Nesting Ecology

1. The response variable is **clutch size** (CSIZE).
1. The covariate is **curved carapace length** (CCL).
1. The factor is **location** (or region).
1. As there are five groups in this analysis and the "Arabian Gulf" turtles are to be the reference group, I created the following four indicator variables.
    * IO=1 if in the "Indian Ocean" group, 0 otherwise
    * RS=1 if in the "Red Sea" group, 0 otherwise
    * CO=1 if in the "Caribbean" group, 0 otherwise
    * WA=1 if in the "West Atlantic" group, 0 otherwise
1. The table connecting the groups to the indicator variable codes is as follows:
```{r echo=FALSE}
data.frame(Region=c("Arabian Gulf","Indian Ocean","Red Sea","Caribbean","West Atlantic"),
           IO=c(0,1,0,0,0),
           RS=c(0,0,1,0,0),
           CO=c(0,0,0,1,0),
           WA=c(0,0,0,0,1)) %>%
  knitr::kable(booktabs=TRUE,row.names=FALSE) %>%
  kableExtra::kable_classic(full_width=FALSE,html_font="Cambria")
```

<ol start="6">
<li>The interaction variables are IO:CCL, RS:CCL, CO:CCL, and WA:CCL.</li>
<ol>

&nbsp;

# Water Quality Near a Gold Mine

1. The response variable is **phosphorous level** measurement (P).
1. The covariate is **distance** from the gold mine (DIST).
1. The factor is **type of phosphorous**.
1. The levels should be changed so that "total" is first; e.g., "total", "dissolved", and "soluble". As there are three groups in this analysis and "total" is to be the reference group, I created the following two indicator variables.
    * DP=1 if the "total dissolved phosphorous" group, 0 otherwise
    * SP=1 if the "soluble reactive phosphorous" group, 0 otherwise
1. The table connecting the groups to the indicator variable codes is as follows:
```{r echo=FALSE}
data.frame(Type=c("Total P","Total Dissolved P","Soluble Reactive P"),
           DP=c(0,1,0),
           SP=c(0,0,1)) %>%
  knitr::kable(booktabs=TRUE,row.names=FALSE) %>%
  kableExtra::kable_classic(full_width=FALSE,html_font="Cambria")
```

<ol start="7">
<li>The interaction variables are DP:DISTANCE and SP:DISTANCE.</li>
<ol>
