---
title: "Exercise Key"
subtitle: "IVR Testing"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/ce/aaaKeys/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_IVRTesting_CE1")
```

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(NCStats)
source("../../../rhelpers/knitr_setup.R")
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
  <li>Use complete sentences to answer questions.</li>
</ul>
</div>

----

# Nurse Wages

1. The hypotheses for the parallel lines test are:
    * H<sub>A</sub> $\mu_{WAGE|EXPER,MALE} = \alpha+\beta EXPER+\delta_{1}MALE+\gamma_{1}EXPER:MALE$.
    * H<sub>0</sub> $\mu_{WAGE|EXPER,MALE} = \alpha+\beta EXPER+\delta_{1}MALE$.
1. The hypotheses for the coincident lines test are:
    * H<sub>A</sub> $\mu_{WAGE|EXPER,MALE} = \alpha+\beta EXPER+\delta_{1}MALE$.
    * H<sub>0</sub> $\mu_{WAGE|EXPER,MALE} = \alpha+\beta EXPER$.
1. The hypotheses for the relationship test are:
    * H<sub>A</sub> $\mu_{WAGE|EXPER,MALE} = \alpha+\beta EXPER$.
    * H<sub>0</sub> $\mu_{WAGE|EXPER,MALE} = \alpha$.
 
&nbsp;

# Turtle Nesting Ecology
```{r TurtleData, echo=FALSE, results="hide"}
ht <- read.csv("https://raw.githubusercontent.com/droglenc/NCData/master/HawksbillTurtles.csv")
ht$Region <- factor(ht$Region,
                    levels=c("Arabian Gulf","Indian Ocean","Red Sea",
                             "Caribbean","West Atlantic"))
ivr.ht <- lm(Clutch.Size~CCL+Region+CCL:Region,data=ht)
anova(ivr.ht)
```
```{r echo=FALSE}
aov.ht <- anova(ivr.ht)
aov.ht.plt <- kPvalue(aov.ht$"Pr(>F)"[3],latex=FALSE)
aov.ht.clt <- kPvalue(aov.ht$"Pr(>F)"[2],latex=FALSE)
aov.ht.rt <- kPvalue(aov.ht$"Pr(>F)"[1],latex=FALSE)
```

1. The models for parallel lines test are as follows:
    * H<sub>A</sub>: $\mu_{CSIZE|CCL,\cdots} = \alpha+\beta CCL+$$\delta_{1}IO+\delta_{2}RS+\delta_{3}+\delta_{4}WA+$ $\gamma_{1}CCL:IO+\gamma_{2}CCL:RS+\gamma_{3}CCL:CO+\gamma_{4}CCL:WA$.
    * H<sub>0</sub>: $\mu_{CSIZE|CCL,\cdots} = \alpha+\beta CCL+$$\delta_{1}IO+\delta_{2}RS+\delta_{3}+\delta_{4}WA$.
1. The models for coincident lines test are as follows:
    * H<sub>A</sub>: $\mu_{CSIZE|CCL,\cdots} = \alpha+\beta CCL+$$\delta_{1}IO+\delta_{2}RS+\delta_{3}+\delta_{4}WA$.
    * H<sub>0</sub>: $\mu_{CSIZE|CCL,\cdots} = \alpha+\beta CCL$.
1. The lines appear to be parallel across all regions (`r aov.ht.plt`). Thus, the relationship between clutch size and curved carapace length does NOT differ among the regions.
1. The intercepts appear to differ among some of the regions (`r aov.ht.clt`). Thus, the mean clutch size at the same curved carapace length, no matter what that carapace length is, differs among some regions.
1. There does appear to be a significant relationship between clutch size and curved carapace length (`r aov.ht.rt`).

#### R Code and Results
```{r prompt=FALSE}
<<TurtleData>>
```

&nbsp;

# Water Quality Near a Gold Mine
```{r GoldMineData, echo=FALSE}
gm <- read.csv("http://derekogle.com/NCMTH207/modules/ce/data/GoldMine.csv")
gm$type <- factor(gm$type,levels=c("total","dissolved","soluble"))
ivr.gm <- lm(phosp~distance+type+distance:type,data=gm)
anova(ivr.gm)
```
```{r echo=FALSE}
aov.gm <- anova(ivr.gm)
aov.gm.plt <- kPvalue(aov.gm$"Pr(>F)"[3],latex=FALSE)
aov.gm.clt <- kPvalue(aov.gm$"Pr(>F)"[2],latex=FALSE)
aov.gm.rt <- kPvalue(aov.gm$"Pr(>F)"[1],latex=FALSE)
```

1. The models for parallel lines test are as follows:
    * H<sub>A</sub>: $\mu_{P|DIST,\cdots} = \alpha+\beta DIST+$$\delta_{1}DP+\delta_{2}SP+$ $\gamma_{1}DIST:DP+\gamma_{2}DIST:SP$.
    * H<sub>0</sub: $\mu_{P|DIST,\cdots} = \alpha+\beta DIST+$$\delta_{1}DP+\delta_{2}SP$.
1. The models for coincident lines test are as follows:
    * H<sub>A</sub>: $\mu_{P|DIST,\cdots} = \alpha+\beta DIST+$$\delta_{1}DP+\delta_{2}SP$.
    * H<sub>0</sub>: $\mu_{P|DIST,\cdots} = \alpha+\beta DIST$.
1. The lines appear to be parallel across all three types of phosphorous (`r aov.gm.plt`). Thus, the relationship between phosphorous level and distance from the gold mine does NOT differ among phosphorous types.
1. The intercepts appear to differ among some of the phosphorous types (`r aov.gm.clt`). Thus, the mean phosphorous level at the same distance from the gold mine, no matter what that distance is, differs among phosphorous types
1. There does appear to be a significant relationship between phosphorous level and distance from the gold mine (`r aov.gm.rt`).

#### R Code and Results
```{r prompt=FALSE, fig.width=4.5}
<<GoldMineData>>
ggplot(data=gm,mapping=aes(x=distance,y=phosp,color=type)) +  
  labs(x="Distance from Gold Mine (km)",y="Phosphrous Level (ppm)") +  
  theme_NCStats() +  
  geom_smooth(method="lm",se=FALSE)
```
