---
title: "Exercise Key"
subtitle: "IVR Variables"
author: Derek H. Ogle
layout: page
css: "/css/modules.css"
output:
  html_document:
    fig_height: 3.5
    fig_width: 3.5
    self_contained: false
---

```{r echo=FALSE, eval=FALSE, warning=FALSE}
## Renders an appropriate HTML file and moves to CE directory
source("modules/ce/aaaKeys/zzz_modHTML_CEKeys.R")
modHTML_CEKeys("KEY_IVRVariables_CE1")
```

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(NCStats)
source("../../../rhelpers/knitr_setup.R")
```

----

<div class="alert alert-warning">
<strong>Note:</strong> 
<ul>
  <li>Use complete sentences to answer questions.</li>
</ul>
</div>

----

# Nurse Wages

1. The response variable is wages.
1. The covariate is months of experience.
1. The factor is gender (female and male).
1. As there are two groups in this analysis and females comes before males, I created the following indicator variable.
    * FEM=1 if "female", 0 otherwise (i.e., "male")
1. The very simple table connecting the groups to the indicator variables codes is as follows:
```{r echo=FALSE}
data.frame(group=c("female","male"),FEM=c(0,1)) %>%
  knitr::kable(booktabs=TRUE,row.names=FALSE) %>%
  kableExtra::kable_classic(full_width=FALSE,html_font="Cambria")
```

<ol start="6">
<li>The only interaction variables would be between months of experience and FEM; thus, experience:FEM.</li>
<ol>

&nbsp;

# Turtle Nesting Ecology

1. The response variable is clutch size.
1. The covariate is curved carapace length (CCL).
1. The factor is location (or region).
1. As there are five groups in this analysis and the "Arabian Gulf" turtles are to be the reference group, I created the following four indicator variables.
    * IO=1 if in the "Indian Ocean" group, 0 otherwise
    * RS=1 if in the "Red Sea" group, 0 otherwise
    * CO=1 if in the "Caribbean" group, 0 otherwise
    * WA=1 if in the "West Atlantic" group, 0 otherwise
1. The table connecting the groups to the indicator variable codes is as follows:
```{r echo=FALSE}
data.frame(Region=c("Arabian Gulf","Indian Ocean","Red Sea","Caribbean","West Atlantic"),
           IO=c(0,1,0,0,0),
           RS=c(0,0,1,0,0),
           CO=c(0,0,0,1,0),
           WA=c(0,0,0,0,1)) %>%
  knitr::kable(booktabs=TRUE,row.names=FALSE) %>%
  kableExtra::kable_classic(full_width=FALSE,html_font="Cambria")
```

<ol start="6">
<li>The interaction variables are IO:CCL, RS:CCL, CO:CCL, and WA:CCL.</li>
<ol>

&nbsp;

# Water Quality Near a Gold Mine

1. The response variable is phosphorous level measurement.
1. The covariate is distance from the gold mine (DISTANCE).
1. The factor is type of phosphorous.
1. The levels should be changed so that "total" is first; e.g., "total", "dissolved", and "soluble".
1. As there are three groups in this analysis and "total" is to be the reference group, I created the following two indicator variables.
    * DP=1 if the "total dissolved phosphorous" group, 0 otherwise
    * SP=1 if the "soluble reactive phosphorous" group, 0 otherwise
1. The table connecting the groups to the indicator variable codes is as follows:
```{r echo=FALSE}
data.frame(Type=c("Total P","Total Dissolved P","Soluble Reactive P"),
           DP=c(0,1,0),
           SP=c(0,0,1)) %>%
  knitr::kable(booktabs=TRUE,row.names=FALSE) %>%
  kableExtra::kable_classic(full_width=FALSE,html_font="Cambria")
```

<ol start="7">
<li>The interaction variables are DP:DISTANCE and SP:DISTANCE.</li>
<ol>
