---
layout: page
title: Exercise
subtitle: 2-Way ANOVA
author: Derek H. Ogle
css: /css/modules.css
output:
  html_document:
    self_contained: no
---

```{r echo=FALSE, eval=FALSE}
# Renders HTML file for webpage ... however, a couple of wayward <div>s, <hr>s,
# and <p>s (last two related to HW format note) should be manually deleted from
# the rendered HTML page.
setwd("C:/aaaWork/Web/GitHub/NCMTH207/modules/Anova-2Way/")
source("../../rhelpers/rhelpers.R")
modHTML("Ex_AOV2Plots")
```

----

<div class="alert alert-warning">
  <strong>Note</strong> Your answers to the questions below should follow the <a href="../../resources/hwformat" target="_blank">expectations for homework found here</a>. Due date is on the <a href="../../resources/Dates-Current" target="_blank">Dates page.</a>
</div>

----

## Visualizing Main and Interaction Effects
For each horizontal pair of graphs below determine which effects (i.e., row factor main, column factor main, or interaction effect) are illustrated. Provide a short explanation for each of your answers (i.e., how did you come to your conclusions).

<div class="alert alert-info">
  <strong>Note</strong> For each horizontal pair of graphs there are three levels of each factor -- some levels may appear to be missing because they overlap with other levels. Make sure to make an explicit statement of an effect or no effect for each main and the interaction effect.
</div>

```{r TWAHWSetup, echo=FALSE, results='hide',message=FALSE}
library(FSA)
rf <- rep(c("A","B","C"),times=3)
cf <- rep(c("I","II","III"),each=3)
r1 <- c(3.5,2.5,1.5,3,2,1,2,1,0)
r2 <- rep(c(3.5,1,3.5),times=3)
r3 <- c(3.5,1,3.5,1,3.5,1,3.5,1,3.5)
ylbl <- "Mean of Response"
rlbl <- "Row Factor"
clbl <- "Column Factor"
ylmt <- c(-1,5)
pchs <- 21:23
bg <- col2rgbt("gray20",1/3)
lwd <- 1
```

<ol>
<li>First set.</li>
```{r TWAHW1, echo=FALSE, results='hide', fig.show='hold', fig.width=3.5, fig.height=3.5}
par(mar=c(3.5,3.5,1,1),mgp=c(2,0.2,0),las=1,tcl=-0.2,yaxt="n",cex=1.1)
interaction.plot(rf,cf,r1,type="o",pch=pchs,bg=bg,lwd=lwd,legend=FALSE,
                 ylab=ylbl,xlab=rlbl,ylim=ylmt)
interaction.plot(cf,rf,r1,type="o",pch=pchs,bg=bg,lwd=lwd,legend=FALSE,
                 ylab=ylbl,xlab=clbl,ylim=ylmt)
```

<br><br>
<li>Second set.</li>
```{r TWAHW2, echo=FALSE, results='hide', fig.show='hold', fig.width=3.5, fig.height=3.5}
par(mar=c(3.5,3.5,1,1),mgp=c(2,0.2,0),las=1,tcl=-0.2,yaxt="n",cex=1.1)
interaction.plot(rf,cf,r2,type="o",pch=pchs,bg=bg,lwd=lwd,legend=FALSE,
                 ylab=ylbl,xlab=rlbl,ylim=ylmt)
interaction.plot(cf,rf,r2,type="o",pch=pchs,bg=bg,lwd=lwd,legend=FALSE,
                 ylab=ylbl,xlab=clbl,ylim=ylmt)
```

<br><br>
<li>Third set.</li>
```{r TWAHW3, echo=FALSE, results='hide', fig.show='hold', fig.width=3.5, fig.height=3.5}
par(mar=c(3.5,3.5,1,1),mgp=c(2,0.2,0),las=1,tcl=-0.2,yaxt="n",cex=1.1)
interaction.plot(rf,cf,r3,type="o",pch=pchs,bg=bg,lwd=lwd,legend=FALSE,
                 ylab=ylbl,xlab=rlbl,ylim=ylmt)
interaction.plot(cf,rf,r3,type="o",pch=pchs,bg=bg,lwd=lwd,legend=FALSE,
                 ylab=ylbl,xlab=clbl,ylim=ylmt)
```

</ol>
